---
title: '(WIP) Tidy Tuesday: Mario Kart Records'
author: Matthew Henderson
date: '2021-05-28'
slug: mario-kart-records
categories:
  - dataviz
tags:
  - tidy-tuesday
draft: yes
references:
- id: cairoFunctionalArtIntroduction2012
  author:
    - family: Cairo
      given: Alberto
  ISBN: 978-0-13-304136-1
  issued:
    - year: 2012
      month: 8
      day: 22
  language: en
  number-of-pages: '402'
  publisher: New Riders
  source: Google Books
  title: >-
    visualization
    The Functional Art: An introduction to information graphics and
  title-short: The Functional Art
  type: book
editor_options: 
  chunk_output_type: console
---

```{r, include=FALSE}
knitr::opts_chunk$set(
       cache = FALSE,
        echo = TRUE,
     warning = FALSE,
     message = FALSE,
       error = FALSE,
    fig.path = "figure/",
  cache.path = "cache/",
     comment = "#>",
    collapse = TRUE,
   fig.width = 8
)

library(readr)

if(!file.exists('records.csv')) {
  download.file('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-05-25/records.csv', destfile = 'records.csv')
}

if(!file.exists('drivers.csv')) {
  download.file('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2021/2021-05-25/drivers.csv', destfile = 'drivers.csv')
}

records <- read_csv('records.csv')
drivers <- read_csv('drivers.csv')
```


This week's Tidy Tuesday was all about world records
for Mario Kart 64.

The data comes from
[Mario Kart World Records](https://mkwrs.com/mk64/)

There were two data sets,
one for the records themselves

```{r, echo=FALSE}
library(reactable)

reactable(records)
```

and another giving data
about the record holders
and the length of time they held records for.

```{r, echo=TRUE}
reactable(drivers)
```


I didn't look at the
drivers data set at all
and decided to try visualising
the progression of records over time instead.

Lately, I've been reading
"The Functional Art"
by Alberto Cairo
[-@cairoFunctionalArtIntroduction2012]
and wanted to start making
an effort in my own visualisations
to learn some of the lessons
from reading that book.

I was also thinking about
my failure in previous weeks
to post my work with alt-text
on Twitter
and the embarrasment
of reading tweets
```{r echo=FALSE}
blogdown::shortcode('tweet', '1395025771668938757')
```
about that bad habit.
So I was keen
to make amends
this week.

Another inspiration was
https://jack-davison.github.io/posts/2021-05-25-exploring-fun-questions-tidytuesday-2021-week-22-mario-kart-64/
He probably did a better
job with a simpler approach.

I started experimenting
with plotting progression
of records over time
and started thinking
that maybe it's not
a good idea to plot
records for different tracks
on the same plot,
even if that plot is faceted
by track.
There is quite a range
of times in world records
[perhaps R code here]
and maybe it would
be hard to answer
questions about which
records are progressing
faster than others
without a common scale.

So I decided to calculate
an improvement percentage
instead and plot that.
At first it seemed
that I should plot percentage
improvement over the previous
record but those plots
were hard to interpret
because the improvement percentages
went up and down.
So I switched to plotting percentage
improvement compared to the first record.
[show how to calculate here]
These plots look
how you would expect
improvements in world records to progress.
They go up in little discrete steps,
sometimes with bigger jumps
and only ever improve.
I thought they were fairly easy
to compare.

Early I noticed that
there was a significant difference
in the records involving shortcuts.
It seems that those world records
can suddenly improve very quickly.
I assumed this was due to new
shortcuts being discovered
but never got around to confirming that fact.

Eventually I settled
on a visualisation that would
compare the effect of shortcuts
on world records.

My final plot is a side-by-side
comparison between the improvements
of world records when no shortcut
is used
and when shortcuts are used.
I chose to look only at
records based on complete races
because the shortcut effect is
only really noticed in those cases.
I think one of the weaknesses
of the final plot is that
it's not immediately obvious
that the comparison is
between shortcut
and no shortcut
or what that means.
I probably should have
made that much clearer
with some expository text
and perhaps with larger
headings on facets.

I think a strength of the
plot is using colour to
emphasise just a few tracks,
those where the shortcuts
has the greatest effect.
I think using the overlaying
of progressions makes
it easy to see that in
the non shortcut case
that world records for
most tracks progress in a
similar way,
presumably proportional
to the initial world record.
I think uses facets in
this case would have made
it more difficult
to see that,
especially if different
axes scales are used
for different tracks.
I also found that trying
to colour every series
differently just confused
the overlaid series.
So a good use of colour,
I think,
was to emphasise just
a few tracks
and the few I picked
were those where the
shortcuts had the greatest effect.

I wasn't really happy
with the colouring I chose.
I was keen not to
try and use accessible
colours for the plot
background and text.
As with the colour
choices for this blog
I used:
http://web-accessibility.carnegiemuseums.org/design/color/

I don't really know
very much about how
to choose colours
for plots in a more
accessible way.
In the end I chose
the RColorBrewer
Dark2 pallette.
In the past I would
have made the lines
thinner and used more alpha
to enhance the beauty
of the overall plot
but I was reluctant
to do that
after learning about
difficulties some people
have seeing colours
when there isn't much contrast between them.

My code is here:
https://github.com/MHenderson/mario-kart-records

![Plot shows two line graphs, side-by-side. Both plots show percentage improvement in world records for Mario Kart 64 over time for multiple tracks. First plot is for world records without using a shortcut. Second plot is for world records using a shortcut. In both plots the same four tracks are highlighted, showing that, for those four tracks, improvements are dramatic (some way over forty percent) and seem to happen quickly, presumably as new shortcuts are discovered. World records without shortcuts, on the other hand, improve slowly and never reach as high as ten percent.](https://github.com/MHenderson/mario-kart-records/raw/master/mario-kart.png)

## References
