---
title: "Extracting Text from Images with Google Vision API and R"
author: Matthew Henderson
slug: handwriting-to-text
date: '2021-11-04'
categories:
  - handwriting-recognition
tags:
  - ocr, computer-vision, gcp, apis, cloud-computing
draft: no
editor_options: 
  chunk_output_type: console
references:
- id: urbanekBase64encToolsBase642015
  author:
    - family: Urbanek
      given: Simon
  genre: manual
  issued:
    - year: 2015
  title: 'base64enc: Tools for base64 encoding'
  type: report
  URL: http://www.rforge.net/base64enc
- id: wickhamHttrToolsWorking2020
  author:
    - family: Wickham
      given: Hadley
  genre: manual
  issued:
    - year: 2020
  title: 'httr: Tools for working with URLs and HTTP'
  type: report
- id: robinsonGutenbergrDownloadProcess2020
  author:
    - family: Robinson
      given: David
  genre: manual
  issued:
    - year: 2020
  title: 'gutenbergr: Download and process public domain works from project gutenberg'
  type: report
- id: oomsJsonlitePackagePractical2014
  author:
    - family: Ooms
      given: Jeroen
  container-title: arXiv:1403.2805 [stat.CO]
  issued:
    - year: 2014
  title: >-
    and r objects
    The jsonlite package: A practical and consistent mapping between JSON data
  type: article-journal
  URL: https://arxiv.org/abs/1403.2805
- id: derlooStringdistPackageApproximate2014
  author:
    - family: Loo
      given: M.P.J.
      non-dropping-particle: der
      dropping-particle: van
  container-title: The R Journal
  issue: '1'
  issued:
    - year: 2014
  page: 111-122
  title: The stringdist package for approximate string matching
  type: article-journal
  URL: https://CRAN.R-project.org/package=stringdist
  volume: '6'
- id: wickhamUsethisAutomatePackage2021
  author:
    - family: Wickham
      given: Hadley
    - family: Bryan
      given: Jennifer
  genre: manual
  issued:
    - year: 2021
  title: 'usethis: Automate package and project setup'
  type: report
- id: oomsTesseractOpenSource2021
  author:
    - family: Ooms
      given: Jeroen
  genre: manual
  issued:
    - year: 2021
  title: 'tesseract: Open source OCR engine'
  type: report
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  cache = TRUE,
  echo = TRUE,
  warning = FALSE,
  message = FALSE,
  error = FALSE,
  fig.path = "figure/",
  cache.path = "cache/",
  comment = "#>",
  collapse = TRUE,
  fig.width = 8
)

ggplot2::theme_set(hrbrthemes::theme_ipsum_rc())

# structure
# 
# 0. introduction
# 1. preparation
# 2. extracting handwriting from an image
# 3. handling multiple pages
# 4. how well does it work?
# 5. Appendix: Google Cloud Platform
```

In this post I'll show you
how to use
Google Cloud Platform's
Vision API
to extract text from images in R.

I'm mostly interested in the case when the
text being extracted is handwritten.
For typewritten or printed text
I use
[Tesseract](https://github.com/tesseract-ocr).

Is it feasible to use
Vision API for this task?
Tesseract works very well
with typewritten or printed text
but does not seem to handle
handwriting as well.
^[Disclaimer: this might just be my lack of knowledge of Tesseract]
Google's Vision API,
on the other hand,
is able to extract handwritten text
from images.

# Preparation

Before doing anything in R
you must
create a project on
Google Cloud Platform (GCP)
and enable Vision API
for your project.

In the appendix
are
some brief notes
about creating
a Vision-enabled
project on GCP.
However,
if you are using GCP
for the first time,
I recommend
following a tutorial
like
[this one](https://cloud.google.com/vision/docs/setup)
instead.

After setting up things
on GCP
you should have
an access token.

To run any of the code below
you must tell R
what the access token is.
But you have to
be careful not to hard-code
the access token
in your source code.
The access token
bestows on its owner
the power to make
API requests
which can incur costs
for the owner
of the GCP project.

To avoid putting
the access token
in your source code
you can
define
an environment variable
`GCLOUD_ACCESS_TOKEN`
to be equal to the access token.
Then you can access the value
of the environment variable
anywhere in your code
with
`Sys.getenv("GCLOUD_ACCESS_TOKEN")`.

While
it is possible
to use `Sys.setenv`
to define an environment variable
a better method
is to define it
in an `.Renviron` file.
One benefit of this approach
is that the environment variable
then persists between sessions.

The
{usethis}
(@wickhamUsethisAutomatePackage2021)
package
makes it easy
to edit
`.Renviron`,
at project
or user scope.

```r
usethis::edit_r_environ(scope = "user")
```
opens
the user
`.Renviron`
in RStudio.
Edit the file
so that it contains
a line like this:

```
GCLOUD_ACCESS_TOKEN=<your access token here>
```

Be careful not to commit this file
to a Git repository.
Anyone with the token
can potentially make requests
of Vision API
through your project
and force you to incur costs.

After defining your access token in
`.Renviron`
you must restart R.
After restarting
you should be able to use
`Sys.getenv`
to check
that the environment variable
is now equal to the access token.

```r
Sys.getenv("GCLOUD_ACCESS_TOKEN")
```

Now you are ready
to begin extracting text from images
in R.

In the next section I'll show you how to
extract text from a single image file.

Later I'll show you how to
extract text from multiple images.

# Extracting handwriting from a single image

In this section
and subsequent sections
I will make use of several images
of my own handwriting.

So you can compare
the output
of Vision API
against a known text
I have handwritten
several pages
of
[James Oliver Curwood](https://en.wikipedia.org/wiki/James_Oliver_Curwood)'s
novel
*Baree, Son of Kazan*.
Later I'll
show you how to compare
the results of
extracting text from these
handwritten pages
to the
[text version of the novel from Project Gutenberg.](https://www.gutenberg.org/ebooks/4748)

## Creating document text detection requests

To use Vision API
to extract text from images
we must
send images
to the URL
`https://vision.googleapis.com/v1/images:annotate`
in JSON format.
Putting images
into JSON
means encoding them
as text.
Vision API requires
images encoded as text
to use the Base64 encoding scheme.
(see
https://en.wikipedia.org/wiki/Base64
for more details).

The JSON payload
of a text detection request
to Vision API
should
look something like this:
```json
{
  "requests": [
    {
      "image": {
        "content": <encoded-image>
      },
      "features": [
        {
          "type": "DOCUMENT_TEXT_DETECTION"
        }
      ]
    }
  ]
}
```
where
`<encoded-image>`
is a Base64 encoding
of the image
we want to extract text from.

Fortunately,
you don't have to figure out
how to do the encoding yourself.
The R package
{base64enc}
(@urbanekBase64encToolsBase642015)
can do that for you.

The
`document_text_detection`
function below
uses
`base64enc::base64encode`
to compute a Base64 encoding
of the image
specified by the input path.
It then packs
the resulting encoding
into a list
of the right format
for converting
to JSON before
being posted
to the URL above.

```{r document_text_detection}
document_text_detection <- function(image_path) {
  list(
    image = list(
      content = base64enc::base64encode(image_path)
    ),
    features = list(
      type = "DOCUMENT_TEXT_DETECTION"
    )
  )
}
```

As you will
be using
{httr}
(@wickhamHttrToolsWorking2020)
you don't even have to worry
about doing the conversion
to JSON yourself.
{httr} will do it
automatically using
{jsonlite}
(@oomsJsonlitePackagePractical2014)
behind-the-scenes
when you make your request.

Now you are ready
to use
`document_text_detection`
to create a request
based on the image
located at:
`~/workspace/baree-handwriting-scans/001.png`

```{r dtd_001}
dtd_001 <- document_text_detection("~/workspace/baree-handwriting-scans/001.png")
```

Be wary of inspecting the return value
of this function call.
It contains a huge base64 string
and R might take a very long time
to output the string
to the screen.

You can use
`substr`
to inspect part of it,
if you really want.

```{r substr_dtd_001}
substr(dtd_001, 1, 200)
```

Another pitfall
to be wary of
is accidentally
encoding the path to a file
instead of the contents of the file itself.
If the response from the
Vision API has error messages
containing paths
then this can be a possible cause.

## Posting requests to Vision API

Now you can use
`httr:POST`
from {httr}
to post your request
to the Vision API.

`httr::POST`
requires at least
`url`
and
`body`
arguments.

`url`
is the webpage
to be retrieved.
In this case
`"https://vision.googleapis.com/v1/images:annotate"`

`body`
is the
request payload.
In this case
a named list
with one element
named `requests`
whose value is a list
of request objects.

As well as those
required arguments
you also have
to tell the server
that the content of your
request is JSON.
You do this
by adding
"Content-Type: application/json"
to the header
of your request.
This means
passing a call
to
`httr::content_type_json()`
as one of the
optional arguments
to
`httr::POST`.

The Vision API documentation
says that for a request
to be accepted
it must have
an
`Authorization = "Bearer <GCLOUD_ACCESS_TOKEN>"`
header.
We can add the header
using the
`httr::add_headers`
function.

```r
httr::add_headers(
    "Authorization" = paste("Bearer", Sys.getenv("GCLOUD_ACCESS_TOKEN"))
)
```

Here we use
`Sys.getenv("GCLOUD_ACCESS_TOKEN")`
to obtain the value
of our access token.

The
`post_vision`
function below
takes a requests list
as input
and returns
the response
from Vision API's
annotation endpoint
as a list.

```{r post_vision}
post_vision <- function(requests) {
  httr::POST(
    url = "https://vision.googleapis.com/v1/images:annotate",
    body = requests,
    encode = "json",
    httr::content_type_json(),
    httr::add_headers(
      "Authorization" = paste("Bearer", Sys.getenv("GCLOUD_ACCESS_TOKEN"))
    )
  )
}
```

Finally,
you are ready
to use
`post_vision`
to make your first request.

`post_vision`
expects a
list of requests
as input.
In this case
you only have one request
`dtd_001`
which you created earlier
by calling
`document_text_detection`
with the path
to your image.
Nevertheless,
you still have to
pack your single
request inside a list.

```{r l_r_001}
l_r_001 <- list(requests = dtd_001)
```

Calling
`post_vision`
now
sends your image
to Vision API
and returns the response.

```{r r_001}
r_001 <- post_vision(l_r_001)
```

You can use
`httr::status_code`
to check that
a valid response
was received.

```{r status_r_001}
httr::status_code(r_001)
```

The value should be 200.

If you get a 401 instead
then inspect
the value of
`r_001`.
If you see
something like this:
```
Response [https://vision.googleapis.com/v1/images:annotate]
  Date: 2021-10-29 09:29
  Status: 401
  Content-Type: application/json; charset=UTF-8
  Size: 634 B
{
  "error": {
    "code": 401,
    "message": "Request had invalid authentication credentials. Expected OAuth 2 acc...
    "status": "UNAUTHENTICATED",
    "details": [
      {
        "@type": "type.googleapis.com/google.rpc.ErrorInfo",
        "reason": "ACCESS_TOKEN_EXPIRED",
        "domain": "googleapis.com",
```
then you might need to rerun
the gcloud tool
(see section 4 of the appendix)
to generate
a new access token.

In the next section
I'll explain
how to get text
out of
a valid response.

## Getting text from the response

If the response
is valid then you
can use
`httr::content`
to extract the content
of the response.

```{r content_001}
content_001 <- httr::content(r_001)
```

The content
contains a lot
of information.
The text is
contained
inside
`responses[[1]]`
as
`fullTextAnnotation$text`.

```{r baree_hw_001}
baree_hw_001 <- content_001$responses[[1]]$fullTextAnnotation$text
cat(baree_hw_001)
```

A lot of the text is readable
but there is
a lot
to do
to fix all of the errors.
And this is only one page!
Maybe things would be better
for someone with better handwriting
than me.

In the next section I'll try
to quantify how close
the text
extracted by Vision API
from the handwritten pages
is to the original text
by measuring
edit distance
to the equivalent
page of text
from Project Gutenberg.

## How well does it work?

To download the original text
from Project Gutenberg
use the
{gutenbergr}
(@robinsonGutenbergrDownloadProcess2020)
package.

```{r baree}
baree <- gutenbergr::gutenberg_download(4748)
```

If you get an error here,
try a different mirror.

From the text of the entire novel
we can extract just the portion
that corresponds to our scanned pages.
In this case
that means lines 96
to 148.

```{r baree_tx}
baree_tx <- paste(baree$text[96:148], collapse = " ")
```

With a bit more
close inspection
we can
find the substring
of the downloaded text
corresponding to the first handwritten page.
In this case
the substring beginning at
the 1st character
and ending at the 1597th.

```{r baree_tx_001}
cat(baree_tx_001 <- stringr::str_sub(baree_tx, 1, 1597))
```

Now using the
{stringdist}
(@derlooStringdistPackageApproximate2014)
package calculate
the edit distance
between the text
extracted from the handwritten page
by Vision API
and the string
extracted from the
Project Gutenberg text.

```{r dist_hw_tx}
stringdist::stringdist(baree_hw_001, baree_tx_001, method = "lv")
```

Apparently we could change
the handwriting based text
to match
the text from Project Gutenberg
with 174 changes.
Quite a lot for one page!
However,
bear in mind
that edit distance is not
quite the same as the number
of changes you
need to make to fix
the document yourself.
You could use spell check tools
to fix some errors quickly
and
use search-and-replace
to remove systematic errors
like non-alphabetic characters
(assuming your text is entirely alphabetic)
or additional spacing.

In calculating edit distance
we used the default
optimal string alignment method
from
{stringdist}.
But the result is the same
if we use different methods like
Levenshtein distance
(`method = "lv"`)
or
Full Damerau-Levenshtein distance
(`method = "dl"`).

# Handling multiple pages

It is possible to send a multi page document
to Vision API.
A PDF,
for example.
A different approach
to multi page documents
is to send one request
with multiple images,
one image per page.

## Building a request based on multiple images

Begin by putting
images of all pages
to be converted into the same folder.
Here I've used
the folder
`~/workspace/baree-handwriting-scans`.

```{r scans_folder}
scans_folder <- "~/workspace/baree-handwriting-scans"
```

Now use
`list.files`
with
`full.names = TRUE`
and
`pattern = '*.png'`
(assuming your images are PNG format)
to get a list
of all images.

```{r scans}
scans <- list.files(scans_folder, pattern = '*.png', full.names = TRUE)
```

Next,
iterate over
`scans`
with
`purrr::map`
and
`document_text_detection`
to create a
list of
JSON request objects,
one for each page.

```{r scans_dtd}
scans_dtd <- purrr::map(scans, document_text_detection)
```

As before,
wrap
this list of requests
in another list.

```{r l_r}
l_r <- list(requests = scans_dtd)
```

Before calling
`post_vision`
and sending all
images to Vision API.

```{r response}
response <- post_vision(l_r)
```

Notice that
even though there are
multiple images
we still only make one request.
The JSON payload
contains encodings
of all images.

Depending on the
number of images
in the
`requests`
list
the above call to
`post_vision`
may take a few seconds
to return a response.

As before,
check that
a valid response
was received
before opening it up
and looking at what is inside.

```{r status_response}
httr::status_code(response)
```

If the response is valid
you can iterate
through the
`responses`
list
inside the response content
extracting
`fullTextAnnotation`
from each element.

This can be done with
`purrr::map`
by passing the name
`"fullTextAnnotation"`
as the function argument.

```{r responses_annotations}
responses_annotations <- purrr::map(httr::content(response)$responses, "fullTextAnnotation")
```

`purrr::map`
converts
`"fullTextAnnotation"`
into an extractor function
which pulls out
elements of the list argument
having that name.

Using the same feature
of
`purrr::map`
you can reach inside
the annotations
and pull out
any
`text`
elements.
This time using
`purrr::map_chr`
instead of
`purrr:map`
because the output
should be a string.

```{r responses_text}
purrr::map_chr(responses_annotations, "text")
```

The
`folder_to_chr`
function below
puts all the above steps together.

Input to
`folder_to_chr`
is a path
to a folder of images
whose text we want to extract.
The return value
is list of strings
containing all text
contained in those images.

```{r folder_to_chr}
folder_to_chr <- function(scans_folder) {
  scans <- list.files(scans_folder, pattern = "*.png", full.names = TRUE)
  response <- post_vision(list(requests = purrr::map(scans, document_text_detection)))
  responses_annotations <- purrr::map(httr::content(response)$responses, "fullTextAnnotation")
  purrr::map_chr(responses_annotations, "text")
}
```

For example,
to convert all the pages
in
`scans_folder`
and combine the results
into a single string
do something like this:

```{r baree_hw}
baree_hw <- paste(folder_to_chr(scans_folder), collapse = " ")
```

## How well does it work?

As before you can use
the edit distance
between
`baree_hw`
and
the equivalent pages
of the original text.
Here,
that text has been
extracted before
into
`baree_tx`.

```{r vision_performance}
stringdist::stringdist(baree_hw, baree_tx)
```

Is this high?
It's hard to know
without something
to compare against.

For example,
how does it compare
to the edit distance
between a random string
of the same length
as the target text
and the target text itself?

```{r random_text}
random_text <- paste(sample(c(letters, " "), stringr::str_length(baree_tx), replace = TRUE), collapse = "")
stringdist::stringdist(random_text, baree_tx)
```

It's hardly surprising that this is
a much higher number.

Another comparison we could make
is against the result
of using Tesseract
to extract text
from the same handwritten pages.

Fortunately,
the
{tesseract}
(@oomsTesseractOpenSource2021)
package makes this very easy.

```{r folder_to_chr_ts}
folder_to_chr_ts <- function(scans_folder) {
  eng <- tesseract::tesseract("eng")
  scans <- list.files(scans_folder, pattern = "*.png", full.names = TRUE)
  paste(purrr::map(scans, tesseract::ocr, engine = eng), collapse = " ")
}

baree_hw_ts <- folder_to_chr_ts(scans_folder)
```

```{r dist_hw_ts_tx}
stringdist::stringdist(baree_hw_ts, baree_tx)
```

So Vision API does much better than Tesseract
according to this particular little test.
But this is probably not a fair comparison.
Tesseract doesn't claim to be able to
extract text from images of handwriting.
Furthermore,
it may be possible to configure
Tesseract in a way that improves
the results
or to modify the input images
to make them better suited to Tesseract.
Also, I'm only looking at my own handwriting.
It might be that another person's
handwriting works better with Tesseract
than mine.

My objective was just to see
if using Vision API made it possible
to convert my own handwritten documents
into text.
While promising it seems
that I might
spend longer fixing errors
in the resulting text 
than it would take
to type them out from scratch.

# Appendix (Google Cloud Platform)

If you haven't worked with Vision API before
the read the following tutorial first:
https://cloud.google.com/vision/docs/setup

  * You should only need to do steps 0 - 2 once.
  * Step 3 should only need to be done once 
   unless the private key has expired.
  * If you have already
   created a project
   and service account
   jump to step 3. 
  * If you already have the private key as well
   jump to step 4.

## 0. Install the cloud SDK

You need to install
the
`glcoud`
tool
to configure authentication.

The installation instructions are here:
https://cloud.google.com/sdk/docs/install

## 1. Create project

First you need to create a GCP project and enable
Vision API for that project.

You will have to setup billing when creating
the project if you haven't done so before.

  * Go to the Project Selector: 
    https://console.cloud.google.com/projectselector2/home/dashboard
    and select *Create Project*
  * Give your project a name
    and click *Create*.
  * Eventually the Project Dashboard appears.
    At the time of writing
    somewhere in the bottom-left of the dashboard
    is a *Getting Started* panel
    containing a link named *Explore and enable APIs*.
    Click on it.
  * You will be transported to the API dashboard.
  * At the top you should see
    *+ ENABLE APIs AND SERVICES*.
    Click on that.
  * Now you get a search dialog.
  * Type *vision* and press enter.
  * Select *Cloud Vision API* and on the page that appears
    click the *Enable* button.

## 2. Create service account

Then you need to create a service account.

This is a good point to just give up.

Strictly speaking this isn't necessary
but it does seem to be the most straightforward way
of enable authentication for your project.

  * Click the hamburger in the navigation bar
    to open the sidebar menu.
  * Scroll down to *IAM & Admin*
    and select *Service Accounts*.
  * Click on *Create Service Account*.
  * Give your account a name.
  * Click *Create and continue*

## 3. Download private key

  * In the Service accounts dashboard find the service
    account you created and click on the three dots
    below *Action*.
  * Select *Manage Keys* from the drop-down menu.
  * On the page that open click on the *ADD KEY* button.
  * Choose *Create New Key* from the drop-down menu.
  * Click *Create* on the modal dialog that opens.
  * You will be prompted to save your key.
  * Download your private key
    and remember where you save it
    (the save location is referenced below as `<PATH_TO_PRIVATE_KEY>`).

## 4. Use gcloud tool to get access token

Now you can use `gcloud` to get the access token
required in the above tutorial.

First, define an environment variable
`GOOGLE_APPLICATION_CREDENTIALS`
pointing to the location where you saved
the private key.

`$ export GOOGLE_APPLICATION_CREDENTIALS=<PATH_TO_PRIVATE_KEY>`

Then run

`$ gcloud auth application-default print-access-token`

The output will be your access token followed by lots of dots.

# References
